### 2021-05-24 concurrency/docs/theory/theory.tex
add frac & counting logic examples
rwlock hoare proofs

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{dashbox}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{stackengine}
\usepackage{stmaryrd}
\usepackage{wasysym}
\usepackage[]{stix}
\usepackage{graphicx}

@@

\newcommand{\hoare}[3]{\ensuremath{\{ #1 \}~ #2 ~\{ #3 \}}}
\newcommand{\viewshift}{\Rrightarrow}
\newcommand{\unit}{\epsilon}
\newcommand{\true}{\textsf{True}}
\newcommand{\false}{\textsf{False}}
\newcommand{\bool}{\textsf{bool}}
\DeclareMathOperator{\rel}{rel}
\DeclareMathOperator{\TPCM}{TPCM}
\DeclareMathOperator{\MonoidBase}{MonoidBase}
\DeclareMathOperator{\MonoidExt}{MonoidExt}
\DeclareMathOperator{\Frac}{Frac}
\newcommand{\monoid}[1]{\textsf{monoid}~#1}
\renewcommand{\eqdef}{\triangleq}
\newcommand{\borrow}[2]{\textsc{\&}^{#1}#2}
\newcommand{\pointsto}{\hookrightarrow}

@@

Again, we define a ``projection'' function from a view on a refinement TPCM to a base TPCM.
\[
  ProjectView_\delta(v) \eqdef \lambda a .~
      \exists x, y .~ v(x) \land y = Project_\delta(x) \land (x \in domain(rel)) \land y \le a.
\]
For a given $m$, we define $ViewOver(m)$ to be the view of all elements that are at least $m$:
\[
  ViewOver(m) \eqdef \lambda n .~ n \ge m
\]
And we can add two views in the natural way:
\[
  v \otimes w \eqdef \lambda m .~ \exists x, y .~ v(x) \land w(y) \land x \cdot y = m
\]
Finally, we're ready to define the $View_\kappa$ function for a given location, again
recursively, like the $Total_\kappa$ function:
\[
  View_\kappa(\gamma) \eqdef 
      \left(
        \bigotimes_{\kappa' \ltGe \kappa} ViewOver(reserved(\kappa,\gamma))
      \right) \otimes
      \left(
        \bigotimes_{\delta} ProjectView_\delta\left(View_\kappa(\delta \refines_{\rel} \gamma)\right)
      \right)
\]
Now the $View_\kappa$ denotes the ``possible'' states that the total unlive state can be in
for a given $\gamma$, without actually depending on any live state even in the refining locations. The $View_\kappa$ is in some sense used to justify the borrows at $\kappa$, so we will have an invariant,
\begin{align}
  & \forall \kappa, \gamma, m .~
  (\kappa, \gamma, m) \in borrows \implies \forall y .~ View_\kappa(\gamma)(y) \implies m \le y
  \label{invView}
\end{align}
    %\left(
    %    \Bigcircle_{\kappa' \ltGe \kappa} reserved(\kappa,\gamma)
    %  \right) \cdot
    %  \left(
    %    \Bigcircle_{\delta} Project_\delta\left(Total_\kappa(\delta \refines_{\rel} \gamma)\right)
    %  \right)
   %\right)
%\end{align}
(We don't need a condition that $\kappa \ltGe active$, although we won't be able to make use
of the invariant otherwise.)

@@

\textbf{Worked RWLock example}

@@

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{dashbox}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{stackengine}
\usepackage{stmaryrd}
\usepackage{wasysym}
\usepackage[]{stix}
\usepackage{graphicx}
\usepackage{xcolor}

@@

\newcommand{\hoare}[3]{\ensuremath{\{ #1 \}~ #2 ~\{ #3 \}}}
\newcommand{\viewshift}{\Rrightarrow}
\newcommand{\viewshiftBack}{\Lleftarrow}
\newcommand{\unit}{\epsilon}
\newcommand{\true}{\textsf{True}}
\newcommand{\false}{\textsf{False}}
\newcommand{\bool}{\textsf{bool}}
\DeclareMathOperator{\rel}{rel}
\DeclareMathOperator{\TPCM}{TPCM}
\DeclareMathOperator{\MonoidBase}{MonoidBase}
\DeclareMathOperator{\MonoidExt}{MonoidExt}
\DeclareMathOperator{\Frac}{Frac}
\newcommand{\monoid}[1]{\textsf{monoid}~#1}
\renewcommand{\eqdef}{\triangleq}
\newcommand{\borrow}[2]{\textsc{\&}^{#1}#2}
\newcommand{\pointsto}{\hookrightarrow}

@@

Again, we define a ``projection'' function from a view on a refinement TPCM to a base TPCM.
\[
  ProjectView_\delta(v) \eqdef \lambda a .~
      \exists x, y .~ v(x) \land y = Project_\delta(x) \land (x \in domain(rel)) \land y \le a.
\]
For a given $m$, we define $ViewOver(m)$ to be the view of all elements that are at least $m$:
\[
  ViewOver(m) \eqdef \lambda n .~ n \ge m
\]
And we can add two views in the natural way:
\[
  v \otimes w \eqdef \lambda m .~ \exists x, y .~ v(x) \land w(y) \land x \cdot y = m
\]
Finally, we're ready to define the $View_\kappa$ function for a given location, again recursively, like the $Total_\kappa$ function:
\[
  View_\kappa(\gamma) \eqdef 
      \left(
        \bigotimes_{\kappa' \ltGe \kappa} ViewOver(reserved(\kappa',\gamma))
      \right) \otimes
      \left(
        \bigotimes_{\delta} ProjectView_\delta\left(View_\kappa(\delta \refines_{\rel} \gamma)\right)
      \right)
\]
Now the $View_\kappa$ denotes the ``possible'' states that the total unlive state can be in
for a given $\gamma$, without actually depending on any live state even in the refining locations. The $View_\kappa$ is in some sense used to justify the borrows at $\kappa$, so we will have an invariant,
\begin{align}
  & \forall \kappa, \gamma, m .~
  (\kappa, \gamma, m) \in borrows \implies \forall y .~ View_\kappa(\gamma)(y) \implies m \le y
  \label{invView}
\end{align}
    %\left(
    %    \Bigcircle_{\kappa' \ltGe \kappa} reserved(\kappa,\gamma)
    %  \right) \cdot
    %  \left(
    %    \Bigcircle_{\delta} Project_\delta\left(Total_\kappa(\delta \refines_{\rel} \gamma)\right)
    %  \right)
   %\right)
%\end{align}
(We don't need a condition that $\kappa \ltGe active$, although we won't be able to make use
of the invariant otherwise.)

@@

\textbf{Fractional logic}

Given a TPCM $M$, we can construct a TPCM $Frac_M$ with the following laws.

\newcommand{\FFrac}{\textsf{Frac}}

\begin{align*}
  \dboxed{x}^\gamma &\viewshiftBack \viewshift \dboxed{\FFrac(x, 1)}^{\alpha \refines \gamma}\\
  \dboxed{\FFrac(x, r)}^{\alpha \refines \gamma}
    *
  \dboxed{\FFrac(x, q)}^{\alpha \refines \gamma}
    &\viewshiftBack \viewshift
  \dboxed{\FFrac(x, r+q)}^{\alpha \refines \gamma}
\end{align*}
And,
\begin{align*}
  \borrow{\kappa}{\dboxed{\FFrac(x, r)}^{\alpha \refines \gamma}} &\viewshift
  \borrow{\kappa}{\dboxed{x}^\gamma}
\end{align*}

\textbf{Counting logic}

\newcommand{\Counter}{\textsf{Counter}}
\newcommand{\SharedRef}{\textsf{SharedRef}}

\begin{align*}
  \dboxed{x}^\gamma &\viewshiftBack \viewshift \dboxed{\Counter(x, 0)}^{\alpha \refines \gamma}\\
  \dboxed{\Counter(x, n)}^{\alpha \refines \gamma}
    &\viewshiftBack \viewshift
  \dboxed{\Counter(x, n+1)}^{\alpha \refines \gamma}
      * \dboxed{\SharedRef(x)}^{\alpha \refines \gamma}
\end{align*}
And,
\begin{align*}
  \borrow{\kappa}{\dboxed{\SharedRef(x)}^{\alpha \refines \gamma}} &\viewshift
  \borrow{\kappa}{\dboxed{x}^\gamma}
\end{align*}

\textbf{RwLock Logic}

@@

\textbf{RwLock Verified Example}

\newcommand{\rwlock}{rwlock}
\newcommand{\exc}{exc}
\newcommand{\rc}{rc}
\newcommand{\Pointer}{Pointer}

\newcommand{\AcquireExc}{\text{AcquireExc}}
\newcommand{\AcquireShared}{\text{AcquireShared}}
\newcommand{\AcquireExcSetBit}{\text{AcquireExcSetBit}}
\newcommand{\AcquireExcCheckRc}{\text{AcquireExcCheckRc}}
\newcommand{\AcquireSharedIncRc}{\text{AcquireSharedIncRc}}
\newcommand{\AcquireSharedCheckExc}{\text{AcquireSharedCheckExc}}
\newcommand{\AcquireSharedAbort}{\text{AcquireSharedAbort}}
\newcommand{\ReleaseExc}{\text{ReleaseExc}}
\newcommand{\ReleaseShared}{\text{ReleaseShared}}

\newcommand{\langloop}{\textbf{loop}}
\newcommand{\until}{\textbf{until}}
\newcommand{\langif}{\textbf{if}}
\newcommand{\langthen}{\textbf{then}}
\newcommand{\langelse}{\textbf{else}}

\newcommand{\CAS}{\textbf{CAS}}
\newcommand{\AtomicAdd}{\textbf{AtomicAdd}}

\newcommand{\Inv}{Inv}
\newcommand{\inv}{inv}
\newcommand{\inlinehoare}[1]{{\color{blue}\left\{#1\right\}}}

Let's use a struct $\rwlock := \{\exc: \Pointer, \rc: \Pointer\}$. Our implementation will be straightforward:

\begin{align*}
  \AcquireExc(\rwlock) ~\eqdef~
      & \langloop~\until~\AcquireExcSetBit(\rwlock);\\
      & \langloop~\until~\AcquireExcCheckRc(\rwlock)\\
      \\
  \AcquireShared(\rwlock) ~\eqdef~
      & \langloop~\until~(\\
        & ~~~~ (\langloop~\until~!\rwlock.\exc);\\
        & ~~~~ \AcquireSharedIncRc(\rwlock);\\ %\AtomicAdd(\rwlock.\rc, 1);
        & ~~~~ \langif~ \AcquireSharedCheckExc(\rwlock) ~\langthen\\
        & ~~~~~~~~ \true\\
        & ~~~~ \langelse\\
        & ~~~~~~~~ \AcquireSharedAbort(\rwlock); \false\\ %\AtomicAdd(\rwlock.\rc, -1); \false
      )\\
      \\
  \AcquireExcSetBit(\rwlock) ~\eqdef~
      & \CAS(\rwlock.\exc,\false,\true)\\
      \\
  \AcquireExcCheckRc(\rwlock) ~\eqdef~
      & (!\rwlock.\rc) = 0\\
      \\
  \AcquireSharedIncRc(\rwlock) ~\eqdef~
      & \AtomicAdd(\rwlock.\rc, 1)\\
      \\
  \AcquireSharedAbort(\rwlock) ~\eqdef~
      & \AtomicAdd(\rwlock.\rc, -1)\\
      \\
  \AcquireSharedCheckExc(\rwlock) ~\eqdef~
      & (!\rwlock.\exc) = \false\\
      \\
  \ReleaseExc(\rwlock) ~\eqdef~
      & \rwlock.\exc := \false\\
      \\
  \ReleaseShared(\rwlock) ~\eqdef~
      & AtomicAdd(\rwlock.\rc, -1)
\end{align*}
We would like to prove the following specifications, for some predicate $\Inv(\rwlock,\alpha,\gamma,\inv)$. (Here, $\inv$ is the user-defined predicate on the ghost state stored by the $\rwlock$.)

$\AcquireExc$:
\begin{align*}
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) }\\
& \AcquireExc(\rwlock) \\
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) * \dboxed{\ExcGuard}^{\alpha \refines \gamma} * \exists x.~ \dboxed{x}^\gamma \land \inv(x) }
\end{align*}
$\ReleaseExc$:
\begin{align*}
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) * \dboxed{\ExcGuard}^{\alpha \refines \gamma} * \dboxed{x}^\gamma \land \inv(x) }\\
& \ReleaseExc(\rwlock) \\
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) }
\end{align*}
$\AcquireShared$:
\begin{align*}
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) }\\
& \AcquireShared(\rwlock) \\
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) * \exists x.~ \dboxed{\SharedGuard(x)}^{\alpha \refines \gamma} \land \inv(x) }
\end{align*}
$\ReleaseShared$:
\begin{align*}
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) * \dboxed{\SharedGuard(x)}^{\alpha \refines \gamma} \land \inv(x) }\\
& \ReleaseShared(\rwlock) \\
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) }
\end{align*}
To prove these specifications, we will first define $\Inv$:
\begin{align*}
  \Inv(\rwlock) ~\eqdef~
      \exists \iota, \boxed{
        \exists exc,rc,x .~
            \dboxed{\Central(exc,rc,x)}^{\alpha \refines \gamma}
            * (\rwlock.\exc \pointsto exc)
            * (\rwlock.\rc \pointsto rc)
      }^{~\iota}
\end{align*}
Now, we can work out how to verify the above programs.
These will follow from the below:
\begin{align*}
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) }\\
& \AcquireExcSetBit(\rwlock)\\
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) *
    \dboxed{\ExcPending}^{\alpha\refines\gamma} }
\end{align*}
